Using Planning to Generate Natural-Language Sentences:\\ Experiences and Challenges

### 2009-12-18 trunk/Papers/crisp-ff-10/crisp-ff.tex
shorter ff sec

It turns out that, as far as the suitability of commonly used
pre-processing and search techniques are concerned, sentence
generation diverges significantly from most known benchmark
domains. We examine this here, in detail, for the case of FF
\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
details of the pre-processing suddenly become deadly obstacles to
performance, and how certain widely used search techniques -- namely
goal orderings and the relaxed plan heuristic -- are completely off
the mark and partly even detrimental. For some of the observed
pitfalls, we are able to provide simple fixes; others pose serious
challenges to planning research. We focus first on the pre-processing,
then on the search techniques.

@@

FF's pre-process contains two details that, for a long time until they
were detected, were so detrimental in our domain that we were not able
to seriously propose FF as a practical solution. Simple fixes to these
details are so effective that FF is now a practical (if not perfect)
possibility.

First, while pre-processing the ADL formulas, FF has a sub-routine
that checks the instantiated formulas -- quantifiers already compiled
into grounded conjunctions and disjunctions -- for tautologies. This
involves a loop checking all pairs of sub-formulas within every
conjunction and disjunction, testing whether they are
identical/identical modulo their sign. For example, if we find $\phi$
and $\neg \phi$ within a disjunction, then the disjunction is replaced
by TRUE. Now, we haven't tested thoroughly whether this sub-routine
has any notable effect on the formulas in other benchmarks. It
certainly has no such effect in our domain. The effect it {\em does}
have is take a huge amount of runtime on the long goal conjunctions
stating that we do not wish to have any open substitution nodes nor
any distractors. The simple fix is hence to switch this sub-routine
off. In small instances, this gives advantages like $0.29s$ vs.\ $11s$
pre-processing time; in a larger instance we ran, original FF's
pre-process doesn't finish within 15 minutes whereas our fix yields a
pre-processing time of about $3s$.

Second, FF's pre-process contains a very small imprudence that
apparently didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition). The easy ones are instantiated in a
spezialized very effectively implemented procedure, while the hard
ones go through a much less effectively implemented more general
procedure. In particular, that latter procedure uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, there actually are no hard
operators in our domain. But FF builds the data structure anyway, even
though it is not needed. In small instances, this results in a couple
of seconds runtime overhead. In larger instances, it exhausts memory
and causes FF to terminate with a segmentation fault.

@@

First, we consider Koehler and Hoffmann's
\shortcite{koehler:hoffmann:jair-00} ``reasonable goal orderings'',
variants and extensions of which are used in several planning systems
today, e.g.\ \cite{hoffmann:etal:jair-04,richter:etal:aaai-08}. FF as
used in the 2000 competition implements the ideas from
\cite{koehler:hoffmann:jair-00}, approximating reasonable orders as a
pre-process and then partitioning the goal set into a {\em goal
  agenda}, a sequence of goal subsets. Each subset is posed to the
planner in isolation, ignoring all other goals. The final plan is the
concatenation of the plans for all entries.

@@

While the goal agenda and related techniques have proved quite useful
in many domains, in the sentence generation domain they often are
completely off-target. The most striking point concerns the
interaction between the goal to create a sentence -- $\neg
\mathsf{subst}(S,\mathsf{root})$ in the rabbit example -- and the goal
to express a certain meaning -- $\mathsf{sleep}(e,r_1)$ in the
rabbit example. The goal agenda orders the former before the latter
because it detects that, once a meaning is expressed, the sentence
creation has started and it is not possible to start anew
anymore. Unless the inverse also holds,\footnote{The inverse holds if,
  in the given grammar, the only way to create a meaning is to create
  a full sentence -- i.e., there are no side sentences like ``the
  sleeping rabbit''.} $\mathsf{subst}(S,\mathsf{root})$ is alone in
the first goal agenda entry, i.e., there is no goal stating the actual
meaning the sentence is supposed to have. Consequently, the planner
decides to generate {\em any} sentence. If it happens to be the wrong
sentence, the planner is stuck in a dead-end. While that dead end is
recognized by the relaxed plan heuristic -- i.e., it is easy to see
that the actions needed for achieving the goal cannot be applied
anymore -- certainly this is an entirely unintended (and a bit
amusing) outcome of this technique. 
% Switching it off, we {\bf \dots
%example before/after.}

@@

While the goal agenda meltdown may be more amusing than alarming, we
next report that relaxed plan heuristics are not far from such a
meltdown either. They are fatally confused by the weird lifecycle of
facts encoding substitution nodes and distractors. In instances where
these constructs are important for the construction of a valid
sentence, these facts are required to be false in the goal, and most
of them are initially false. However, any valid plan must make many of
them true at some intermediate time point (e.g.\ distractors arise
depending on the decisions taken by the planner). Hence, from the
perspective of the relaxed plan heuristic, from one state to the next
whole classes of goals suddenly appear ``out of the blue''. These
goals are not accounted for in the preceding state, and hence the
relaxed plan is completely un-informative.

Consider again the rabbit example. In the initial state, the relaxed
plan has length $1$ because we only need to achieve $\neg
\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$, which is
done by the single action ``sleeps''. However, that action has the
effect ``forall (?y - individual) (when (not (= ?y ?xself))
(distractor ?subjnode ?y))'' which introduces new distractors. Those
must be removed, so they form ``new'' goals. In the given case, the
relaxed plan now contains two actions to achieve these goals. In
general (if there are $n$ rabbits, say), the relaxed plan may become
arbitrarily long.
%To give just one concrete example, say we are in the initial state,
%and have not yet committed to a sentence. For certain kinds of
%sentences ({\bf more concrete here???}), the relaxed plan has length
%$1$ because the we only need to choose the desired meaning. However,
%in the state $s$ resulting from this action, the mechanisms of the
%planning task introduce all the sub-goals that are required in order
%to pin-point that meaning (in particular we get a new goal for all the
%distractors that must be removed). Hence the relaxed plan is much
%longer than before -- arbitrarily longer, in general.  
For FF's particular search algorithm, enforced hill-climbing (EHC),
this means that {\em FF performs a blind breadth-first search}: EHC
starts at $h$ value $1$ and thereafter needs to solve the whole
problem in order to find a (goal) state with lower $h$ value.

@@

Can we design heuristics that are better at dealing with the lifecycle
of substitution nodes and distractors? The problem of not noticing
them in the initial state is present for all approaches that are
dominated by $h^+$,
e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}. The
same problem is also present for the inadmissible landmarks heuristic
\cite{richter:etal:aaai-08}, as well as for $h^{\mathsf{cea}}$
\cite{helmert:geffner:icaps-08} and its variant $h^{\mathsf{pcc}}$
\cite{cai:etal:icaps-09}, and hence also for $h^{\mathsf{CG}}$
\cite{helmert:icaps-04}. As far as known heuristics are concerned,
hope thus resides mainly in the $h^m$ family and in abstraction
heuristics. We ran a few tests with an example like the rabbit one,
except that there are 10 distractors. The Graphplan estimate for the
initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
$4$ is far away from the real goal distance $14$. %
%{\bf here I could use
%  differently scaled versions to see whether the distance estimate is
%  constant across the number of distractors; I would think that
%  yes}. 
% .. have tried IPP on NEUES-MODIFIER-SET but get seg faults and shit..
% oh, whatever.
With merge\&shrink \cite{helmert:etal:icaps07}, we obtained the
estimate $11$ for the initial state, which is much better. Still, we
could not find a parameter setting that reduced the search nodes to
less than $5200$, which is not much better than blind search which
took $13800$ nodes. So, all in all, finding better heuristic functions
for this kind of problem remains a challenge.

%{\bf \dots
%  (malte agreed to run a test on modifiers-10-10-).. larger versions
%  would also be useful, to see how the search spaces develop}

@@

The only FF technique that appears to work reasonably well is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful, and it certainly prunes
the set of avaibale actions quite effectively. For example, in the
10-distractor test instance, $16875$ nodes are explored without
helpful actions pruning; with pruning, this reduces to $3078$. For
larger examples, the factor between the two numbers tends to become
larger.

Since EHC is often completely inadequate, and helpful actions are
quite effective, our FF configuration directly runs best-first search
with helpful actions (as opposed to original FF which switches to
best-first with{\em out} helpful actions if EHC fails).

>>>>

%It turns out that, as far as the suitability of commonly used
%pre-processing and search techniques are concerned, sentence
%generation diverges significantly from most known benchmark
%domains. We examine this here, in detail, for the case of FF
%\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
%details of the pre-processing suddenly become deadly obstacles to
%performance, and how certain widely used search techniques -- namely
%goal orderings and the relaxed plan heuristic -- are completely off
%the mark and partly even detrimental. For some of the observed
%pitfalls, we are able to provide simple fixes; others pose serious
%challenges to planning research. We focus first on the pre-processing,
%then on the search techniques.

It turns out that FF's exhibits, in this domain, a number of
detrimental phenomena that were previously unheard of. We now examine
these in some detail.  For some of the observed pitfalls, we are able
to provide simple fixes. Others pose serious challenges to planning
research.

%We focus first on the pre-processing,
%then on the search techniques.

@@

FF's pre-process contains two details that, for a long time until the
writing of this paper, were so detrimental that we were not able to
seriously propose FF as a practical solution.
%Simple fixes
%to these details are so effective that FF is now a practical (if not
%perfect) possibility.

While pre-processing the ADL formulas, FF has a sub-routine that
checks the grounded formulas
% -- quantifiers already compiled into
%grounded conjunctions and disjunctions -- 
for tautologies. This involves a loop checking all pairs of
sub-formulas within every conjunction and disjunction, testing whether
they are identical/identical modulo their sign. For example, if we
find $\phi$ and $\neg \phi$ within a disjunction, then the disjunction
is replaced by TRUE. Now, we haven't tested whether this sub-routine
has a notable effect on the formulas in other benchmarks. It certainly
has no such effect in our domain, other than taking a huge amount of
runtime on the long goal conjunctions stating that we do not wish to
have any open substitution nodes nor any distractors. The fix is to
switch this sub-routine off. This trivial operation alone can render
previously infeasible instances (pre-processing time $>15$ minutes)
harmless (instance solved in $6$ seconds {\em total} time).

%In small instances, this gives advantages like
%$0.29s$ vs.\ $11s$ pre-processing time. In a larger instance we ran,
%original FF's pre-process doesn't finish within 15 minutes whereas our
%fix yields a pre-processing time of about $3s$.

FF's pre-process contains a very small imprudence that apparently
didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition), which go through different instantiation
procedures. The procedure for hard operators uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, we actually have no hard
operators. But FF builds the data structure anyway. This sometimes
results in a couple of seconds runtime overhead; it also sometimes
exhausts memory and causes a segmentation fault.

@@

First, we consider Koehler and Hoffmann's
\shortcite{koehler:hoffmann:jair-00} ``reasonable goal orderings'',
variants and extensions of which are in prominent use today, e.g.\ in
LAMA \cite{richter:etal:aaai-08}. FF as used in the 2000 competition
approximates reasonable orders as a pre-process and then partitions
the goal set into a {\em goal agenda}, a sequence of goal
subsets. Each subset is posed to the planner in isolation. The final
plan is the concatenation of the plans for all entries.

@@

The goal agenda is completely off-target in our domain. The most
striking point concerns the interaction between the goal to create a
sentence ($\neg \mathsf{subst}(S,\mathsf{root})$ in the rabbit
example), and the goal to express a certain meaning
($\mathsf{sleep}(e,r_1)$ in the rabbit example). The goal agenda
orders the former before the latter because it detects that, once a
meaning is expressed, the sentence creation has started and it is not
possible to start anew anymore. Unless the inverse also
holds,\footnote{The inverse holds if, in the given grammar, the only
  way to create a meaning is to create a full sentence -- i.e., there
  are no side sentences like ``the sleeping rabbit''.}
$\mathsf{subst}(S,\mathsf{root})$ is alone in the first goal agenda
entry, i.e., there is no goal stating the actual meaning the sentence
is supposed to have. Consequently, the planner decides to generate
{\em any} sentence. If it happens to be the wrong sentence, the
planner is stuck in a dead-end. 
%While that dead end is recognized by
%the relaxed plan heuristic -- i.e., it is easy to see that the actions
%needed for achieving the goal cannot be applied anymore -- certainly
%this is an entirely unintended (and a bit amusing) outcome of this
%technique.
% Switching it off, we {\bf \dots
%example before/after.}

@@

In our improved FF, the goal agenda is simply switched off. Having
disposed of goal orderings in that way, let us turn our attention to
the relaxed plan heuristic. %
%While the goal agenda meltdown may be more amusing than alarming, we
%next report that relaxed plan heuristics are not far from such a
%meltdown either. They are fatally 
This meets its Waterloo as well, due to the weird lifecycle of facts
encoding substitution nodes and distractors. In instances where these
constructs are important for the construction of a valid sentence,
these facts are required to be false in the goal, and most of them are
initially false. However, any valid plan must make many of them true
at some intermediate time point (e.g.\ distractors arise depending on
the decisions taken by the planner). Hence, from the perspective of
the relaxed plan heuristic, from one state to the next whole classes
of goals suddenly appear ``out of the blue''. 
%These goals are not
%accounted for at all in the preceding state.
%, and hence the relaxed plan is
%completely un-informative.

Consider again the rabbit example. In the initial state, the relaxed
plan has length $1$ because we only need to achieve $\neg
\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$, which is
done by the single action ``sleeps''. However, that action has the
effect ``forall (?y - individual) (when (not (= ?y ?xself))
(distractor ?subjnode ?y))'' that introduces new distractors. Those
must be removed, so they form ``new'' goals. In the given case, the
relaxed plan now contains two actions to achieve these goals. In
general (if there are $n$ rabbits, say), the relaxed plan may become
arbitrarily long.
%To give just one concrete example, say we are in the initial state,
%and have not yet committed to a sentence. For certain kinds of
%sentences ({\bf more concrete here???}), the relaxed plan has length
%$1$ because the we only need to choose the desired meaning. However,
%in the state $s$ resulting from this action, the mechanisms of the
%planning task introduce all the sub-goals that are required in order
%to pin-point that meaning (in particular we get a new goal for all the
%distractors that must be removed). Hence the relaxed plan is much
%longer than before -- arbitrarily longer, in general.  
Given FF's particular search algorithm, enforced hill-climbing (EHC),
this means that {\em FF turns into a blind breadth-first search}: EHC
starts at $h$ value $1$ and thereafter needs to solve the whole
problem in order to find a (goal) state with lower $h$ value.

@@

%Can we design heuristics that are better at dealing with the lifecycle
%of substitution nodes and distractors? The problem of not noticing
%them in the initial state is present for all approaches that are
%dominated by $h^+$,
%e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}. The
%same problem is also present for the inadmissible landmarks heuristic
%\cite{richter:etal:aaai-08}, as well as for $h^{\mathsf{cea}}$
%\cite{helmert:geffner:icaps-08} and its variant $h^{\mathsf{pcc}}$
%\cite{cai:etal:icaps-09}, and hence also for $h^{\mathsf{CG}}$
%\cite{helmert:icaps-04}. As far as known heuristics are concerned,
%hope thus resides mainly in the $h^m$ family and in abstraction
%heuristics. We ran a few tests with an example like the rabbit one,
%except that there are 10 distractors. The Graphplan estimate for the
%initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
%$4$ is far away from the real goal distance $14$. %
%{\bf here I could use
%  differently scaled versions to see whether the distance estimate is
%  constant across the number of distractors; I would think that
%  yes}. 
% .. have tried IPP on NEUES-MODIFIER-SET but get seg faults and shit..
% oh, whatever.
%With merge\&shrink \cite{helmert:etal:icaps07}, we obtained the
%estimate $11$ for the initial state, which is much better. Still, we
%could not find a parameter setting that reduced the search nodes to
%less than $5200$, which is not much better than blind search ($13800$
%nodes). All in all, finding better heuristic functions for this kind
%of problem remains a challenge.

Can we design heuristics that are better at dealing with the lifecycle
of substitution nodes and distractors? It is easy to see that the
problem of not noticing them in the initial state is present for most
known heuristics
\cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
two notable eceptions are the $h^m$ family, and abstraction
heuristics. We ran a few tests with an example like the rabbit one,
except that there are 10 distractors. The Graphplan estimate for the
initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
$4$ is far away from the real goal distance $14$. With merge\&shrink
\cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
initial state, which is much better. Still the search space with that
latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
nodes for blind search) and it is not clear how well the technique
will scale to larger instances and instances with a more complex
grammar. Certainly, interesting challenges remain.

@@

The only FF technique that appears to work reasonably well is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful. As our experiments data
will show, it certainly prunes the set of available actions quite
effectively.

%For example, in the
%10-distractor test instance, $16875$ nodes are explored without
%helpful actions pruning; with pruning, this reduces to $3078$. For
%larger examples, the factor between the two numbers tends to become
%larger.

%Since EHC is often completely inadequate, and helpful actions are
%quite effective,

Given the above observations regarding EHC and helpful actions, our
modified FF directly runs best-first search with helpful actions (as
opposed to original FF which switches to best-first with{\em out}
helpful actions if EHC fails).
